---
title: "Untitled"
output: html_document
date: "2022-11-03"
---

```{r}
library(here) #nice file paths
library(readxl) #read in excel data
library(tidyverse) #collection of packages for data science
library(patchwork) #"stiches" together ggplots
library(magrittr) #pipes
library(broom) #tidy displays
library(ggrepel) #for labelling points on a ggplot
library(lubridate) #helps with date objects
library(bacondecomp) #Goodman Bacon Decomposition
library(did) #Callaway and Sant'Anna estimator for DID
library(staggered) #Sun and Abraham estimator for DID

source("./helper_func_ed.R")
source("./helper_func_ed_sum.R")
source("./helper_func_ed_sum_hte.R")

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

library(r2symbols)
```

# Scenario 7 (4 states, staggered timing, hetergeneous effects)

Scenario 7 is similar to Scenario 5 except now the treatment effect is different for the two treated states, i.e., it is heterogeneous. 

```{r}
# here we also specify the range of the data to be read into R because we don't
# want to include other information included in this sheet
s7 <- read_xlsx(here("data", "scenarios.xlsx"), sheet = "scen9", range = "A1:H81",
                col_types = c("text", "numeric", "text", "numeric", "text", "numeric",
                              "date", "date"))
```

## Visualization of Scenario 7

```{r, echo=F}
s7 %<>% mutate(ever_trt = case_when(state %in% c(3, 4) ~ "treated", 
                                    state %in% c(1, 2) ~ "untreated"))
ggplot(s7, aes(x = time, y = outcome)) + 
  geom_line(aes(col = factor(state))) + 
  geom_point(aes(fill = factor(state), pch = ever_trt)) + 
  labs(title = "Scenario 7 (Hetergeneous)") +
  geom_vline(aes(xintercept = 4.5), lty = 2, col = "#005a32") + 
  geom_vline(aes(xintercept = 11.5), lty = 2, col = "#99000d") +
  scale_fill_manual(values = c("#a6cee3","#1f78b4","#005a32", "#99000d")) +
  scale_color_manual(values = c("#a6cee3","#1f78b4","#005a32", "#99000d")) +
  scale_shape_manual(values = c(23, 21)) + 
  theme_bw()

```

**Visualization of all comparisons made by TWFE regression**

```{r, echo=F}
s7_mean_1_4 <- s7 %>% filter(time < 5) %>%
  group_by(state) %>%
  summarise(mean_1_4 = mean(outcome))

s7_mean_5_20 <- s7 %>% filter(time >= 5) %>%
  group_by(state) %>%
  summarise(mean_5_20 = mean(outcome))

s7_DID1 <- bind_cols(s7_mean_1_4, s7_mean_5_20 %>% select(-state)) %>%
  mutate(time_diff = mean_5_20 - mean_1_4)


s7_mean_1_11 <- s7 %>% filter(time < 12) %>%
  group_by(state) %>%
  summarise(mean_1_11 = mean(outcome))

s7_mean_12_20 <- s7 %>% filter(time >= 12) %>%
  group_by(state) %>%
  summarise(mean_12_20 = mean(outcome))

s7_DID2 <- bind_cols(s7_mean_1_11, 
                     s7_mean_12_20 %>% select(-state)) %>%
  mutate(time_diff2 = mean_12_20 - mean_1_11)

s7_mean_5_11 <- s7 %>% filter(time >= 5 & time < 12) %>%
  group_by(state) %>%
  summarise(mean_5_11 = mean(outcome))

s7_DID3 <- bind_cols(s7_mean_1_4, 
                     s7_mean_5_11 %>% select(-state)) %>%
  mutate(time_diff3 = mean_5_11 - mean_1_4)

s7_DID4 <- bind_cols(s7_mean_5_11, 
                     s7_mean_12_20 %>% select(-state)) %>%
  mutate(time_diff4 = mean_12_20 - mean_5_11)
```

```{r, echo = F, fig.height = 8, fig.width = 10}
s7_comp1 <- ggplot(s7 %>% filter(state %in% c(1, 2, 3)), aes(x = time, y = outcome)) + 
  geom_line(aes(col = factor(state))) + 
  geom_point(aes(fill = factor(state), pch = ever_trt)) + labs(title = "A) Contrast 1 (Treated vs. never-treated)") +
  geom_vline(aes(xintercept = 4.5), lty = 2) +
  scale_color_manual(values = c("#a6cee3","#1f78b4","#005a32")) +
  scale_fill_manual(values = c("#a6cee3","#1f78b4","#005a32")) +
  theme_bw() + 
  scale_x_continuous(limits = c(1, 20)) +
  scale_shape_manual(values = c(23, 21)) + 
  geom_label_repel(data = s7_DID1 %>% filter(state %in% c(1,2,3)), 
                  aes(y = mean_1_4, x = 2.5, label = paste("mean_y:", mean_1_4), col = state), show.legend = F) +
  geom_label_repel(data = s7_DID1 %>% filter(state %in% c(1,2,3)), 
                  aes(y = mean_5_20, x = 12.5, label = paste("mean_y:", mean_5_20), col = state), show.legend = F)
 
s7_comp2 <- ggplot(s7 %>% filter(state %in% c(1, 2, 4)), aes(x = time, y = outcome)) + 
  geom_line(aes(col = factor(state))) +
  geom_point(aes(fill = factor(state),  pch = ever_trt)) + labs(title = "B) Contrast 2 (Treated vs. never-treated)") +
  geom_vline(aes(xintercept = 11.5), lty = 2) +
  scale_color_manual(values = c("#a6cee3","#1f78b4","#99000d")) +
  scale_fill_manual(values = c("#a6cee3","#1f78b4","#99000d")) +
  theme_bw()+ 
  scale_x_continuous(limits = c(1, 20)) +
  scale_shape_manual(values = c(23, 21)) + 
  geom_label_repel(data = s7_DID2 %>% filter(state %in% c(1,2,4)), 
                   aes(y = mean_1_11, x = 6, label = paste("mean_y:", mean_1_11), col = state), show.legend = F) +
  geom_label_repel(data = s7_DID2 %>% filter(state %in% c(1,2,4)), 
                   aes(y = mean_12_20, x = 16, label = paste("mean_y:", mean_12_20), col = state), show.legend = F)

s7_comp4 <- ggplot(s7 %>% filter(state %in% c(3, 4), time >= 5), aes(x = time, y = outcome)) +  
  geom_line(aes(col = factor(state))) + 
  geom_point(aes(fill = factor(state), pch = ever_trt)) + labs(title = "C) Contrast 4 (Later vs. earlier-treated)") +
  geom_vline(aes(xintercept = 11.5), lty = 2) +
  scale_color_manual(values = c("#005a32", "#99000d")) +
  scale_fill_manual(values = c("#005a32", "#99000d")) +
  scale_shape_manual(values = c(23)) + 
  theme_bw()+ 
  scale_x_continuous(limits = c(1, 20)) +
  geom_label_repel(data = s7_DID4 %>% filter(state %in% c(3,4)), 
                   aes(y = mean_5_11, x = 8, label = paste("mean_y:", mean_5_11), col = state), show.legend = F) +
  geom_label_repel(data = s7_DID4 %>% filter(state %in% c(3,4)), 
                   aes(y = mean_12_20, x = 16, label = paste("mean_y:", mean_12_20), col = state), show.legend = F)

s7_comp3 <- ggplot(s7 %>% filter(state %in% c(3, 4), time < 12), aes(x = time, y = outcome)) +   
  geom_line(aes(col = factor(state))) + 
  geom_point(aes(fill = factor(state), pch = ever_trt)) + labs(title = "D) Contrast 3 (Earlier vs. later-treated)") +
  geom_vline(aes(xintercept = 4.5), lty = 2) +
  scale_color_manual(values = c("#005a32", "#99000d")) +
  scale_fill_manual(values = c("#005a32", "#99000d")) +
  scale_shape_manual(values = c(23)) + 
  theme_bw()+ 
  scale_x_continuous(limits = c(1, 20)) +
  geom_label_repel(data = s7_DID3 %>% filter(state %in% c(3,4)), 
                   aes(y = mean_1_4, x = 2.5, label = paste("mean_y:", mean_1_4), col = state), show.legend = F) +
  geom_label_repel(data = s7_DID3 %>% filter(state %in% c(3,4)), 
                   aes(y = mean_5_11, x = 8, label = paste("mean_y:", mean_5_11), col = state), show.legend = F)

s7_comp1 + s7_comp2 + s7_comp3 + s7_comp4 + plot_layout(guides = "collect")
```

**Contrast 1:**

* Pre-post difference for never-treated state 1 is: 43.5-13.5=30
* Pre-post difference for never-treated state 2 is: 44.5-14.5=30
* Implying the average pre-post difference equals 30 for the never-treated states
* Pre-post difference for treated state 3 is 52.5-16.5 = 36
* $DID_{3 vs 1,2} = 36-30 = 6$

**Contrast 2:**

* Pre-post difference for never-treated state 1 is: 54-24=30
* Pre-post difference for never-treated state 2 is: 55-25=30
* Implying the average pre-post difference equals 30 for the never-treated states
* Pre-post difference for treated state 4 is 67-33= 34
* $DID_{4 vs 1,2} = 34-30 = 4$

**Contrast 3:**

* Pre-post difference for later-treated state 4 is: 39-22.5=16.5
* Pre-post difference for earlier-treated state 3 is: 39-16.5=22.5
* $DID_{3 vs 4} = 22.5-16.5 = 6$

**Contrast 4:**

* Pre-post difference for earlier-treated state 3 is: 63-39=24
* Pre-post difference for later-treated state 4 is: 67-39=28
* $DID_{4 vs 3} = 28-24 = 4$

Contrasts 1 and 2 are definitely okay because they are clean comparisons between never-treated and treated states. Contrasts 3 and 4 are trickier, however, we can see from the diagram that parallel trends is still satisfied. So, even though 
the model uses a previously treated state as a control, this is okay because the causal effect of the treatment was heterogeneous and does not lead to a violation of the parallel trends assumption. Thus, we are comfortable with all contrasts 
contributing to the TWFE regression estimate. To see how much weight is put on 
each one, we use the Goodman Bacon decomposition, but first we consider the 
different causal effects that can be calculated.

## Calculation of summary measures of the causal effect

Once the setting includes both heterogeneous effects and staggered timing, there
is a question of how to aggregate the causal effect estimates across multiple
treated units into one estimate. Here are three causal effects that could be 
calculated:

* **Causal Effect A:** Take an average of the estimated effects across the treated units: 

$\frac{\text{Number in adoption cohort 1}}{\text{Number of treated units}} \times ATT_{adoption cohort 1} + \frac{\text{Number in adoption cohort 2}}{\text{Number of treated units}} \times ATT_{adoption cohort 2}$

$\frac{1}{2} \times 6 + \frac{1}{2} \times 4 = 5$

* **Causal Effect B:** Take an average of the effects across all post-treatment
time points:

$\frac{ATT_{t=1} + ATT_{t=2} + ... + ATT_{t=16}}{\text{Number of posttreatment times}}$

$\frac{[5\times9] +[6\times7]}{16}=5.43$

* **Causal Effect C:** Take a weighted average of the post-treatment-time-specific 
effects, where the weights correspond to the number of units treated at each time point:

   $\frac{2\times{[ATT_{t=1}} +{ATT_{t=2}} + ... + {ATT_{t=9}] + 1\times[{ATT_{t=10} + ... +ATT_{t=16}}}]}{\text{Total number of state-time points}}$
   
   $\frac{2\times(5+5+...+5) + 1\times{(6+...+6)}}{25} = 5.28$
   
Each of these causal effects are valid measures to calculate and all come with 
strengths and weaknesses. The researcher needs to decide which one to calculate.
Let's first see how the estimate from the TWFE model compares.

## TWFE Regression Model

```{r}
s7_mod <- lm(outcome ~ policy + state + factor(time), 
             data = s7)
tidy(s7_mod)
```

The coefficient of the policy term equals `r s7_mod$coefficients["policy1"] %>% round(2)`.
This is smaller than any of **Causal Effect A**, **Causal Effect B**, or **Causal Effect C**.
To see how the TWFE came to be this number, we use the Goodman Bacon 
decomposition to see how much weight each of the contrasts contributes to the 
effect estimate.

## Goodman Bacon Decomposition

```{r}
s7 %<>% mutate(state_n = as.numeric(as.character(state)),
               policy_n = as.numeric(as.character(policy)))

s7_bacon <- bacon(outcome ~ policy_n,
      data = s7,
      id_var = "state_n",
      time_var = "time")
```

It is more helpful to view the four contrasts separately:

```{r}
s7_bacon
```

The estimates of the treatment effects for each contrast are as we
calculated them above. However, the TWFE's estimate of 
`r sum(s7_bacon$estimate*s7_bacon$weight) %>% round(2)`,
is pulled towards 4, because the combined weight put on
the effect estimate of 4 is higher than the combined weight put on the effect
estimate of 6. This leads to an effect estimate from TWFE that is smaller than 
any of the causal effect estimates.

The TWFE regression estimate is estimating a different causal estimand than any 
of the three with initially proposed. The new proposed estimators overcome this
issue and are considered next.

## Group-Time ATT Estimator

Like in the earlier examples we start by running the `att_gt()` function. 

```{r, warning=F}
s7_cs <- att_gt(yname = "outcome", 
             tname = "time", 
             idname = "state_n", 
             gname = "time_first_treat", 
             data = s7, 
             control_group = "notyettreated",
             anticipation = 0)
```

Callaway and Sant'Anna provide many options for aggregating the group-time 
effects. The simplest option is specify `type = simple` in the `attge()` function.
This estimate considers only the contrasts with a never-treated state (i.e., 
Comparisons 1 and 2 in the figure above) and combines them into a weighted 
average, where the weights correspond to each adoption cohort's time spent in 
the post-treatment period. For contrast 1, there are 16 time periods post 
treatment and for contrast 2, there are 9 periods post-treatment. Thus the 
weighted average is: $6\times(16/25) + 4\times(9/25)=$ `r 6*(16/25) + 4*(9/25)`.
This is the same as **Causal Effect C** calculated above.

```{r}
#aggregate the group time average treatment effects
#type = "simple": weighted average of all group-time average treatment effects
#with weights proportional to the group size
s7_cs_ag <- aggte(s7_cs, type="simple")
summary(s7_cs_ag)
```

Alternatively, we can specify `type = "group"` to get separate effect estimates 
according to time of implementation. In the output `Group` denotes the time of 
the policy's introduction for the treated states.

```{r}
s7_cs_ag2 <- aggte(s7_cs, type = "group")
summary(s7_cs_ag2)
```

The output also displays an estimate of the Overall ATT, equal to `r s7_cs_ag2$overall.att`.
This estimate is different from the one estimated `where type = "simple"`. Here,
the estimate is a weighted average, with weights proportional to the number of 
units in each adoption cohort (e.g., $(1/2)\times 6 + (1/2) \times 4=$ `r s7_cs_ag2$overall.att`). This is equal to **Causal Effect A** calculated above.

Alternatively, we can specific `type = "dynamic"` to get separate effect estimates
for each time period. We can also see that the Overall ATT from this call is 
equal to **Causal Effect B** calculated above.

```{r}
s7_cs_ag2 <- aggte(s7_cs, type = "dynamic")
summary(s7_cs_ag2)
```

## Target Trial Estimator

`r colorize("Dana, need help choosing which fit_event_* function to use for this one", "green")`

We can also estimate the effects using the Target Trial estimator. This time, we
start with the `fit_event_jack()` function which gives an estimate for each time
since treatment. 

```{r}
s7_bm <- fit_event_jack(outcome_var = "outcome", 
                            date_var = "time_as_date", 
                            unit_var = "state",
                            policy_var = "time_first_trt_date", 
                            data = s7, 
                            max_time_to = 10e7)

s7_bm_ave <- s7_bm %>% filter(cohort == "average")

s7_bm_ave %<>% mutate(lb = estimate - 1.96*se,
                      ub = estimate + 1.96*se)
```

The plot below shows that the time-specific estimates equal 5 when there are
two states contributing to the estimate (one with a causal effect of 4 and the 
other with a causal effect of 6) and 6 when only the early-introduction state 
contributes to the causal effect estimate.

```{r}
ggplot(s7_bm_ave, aes(x = event_time, y = estimate)) + 
  geom_hline(yintercept = 0, lty = 2) +
  geom_point(aes(col = event_time >= 0)) + 
  geom_linerange(aes(ymin = lb, ymax = ub, col = event_time >= 0)) + 
  labs(y = "Estimate", x = "Event time") + 
  theme_bw() + 
  scale_color_discrete(labels=c('pre', 'post')) +
  theme(legend.title=element_blank(), legend.position = "bottom")
```

To summarize the effects into one aggregated effect estimate
we use the fit_event_jack_sum() function:

```{r}
s7_bm2 <- fit_event_jack_sum(outcome_var = "outcome", 
                            date_var = "time_as_date", 
                            unit_var = "state",
                            policy_var = "time_first_trt_date", 
                            data = s7, 
                            max_time_to = 10e7)

s7_bm2$estimate
```

```{r}
s7_bm3 <- fit_event_jack_sum_hte(outcome_var = "outcome", 
                            date_var = "time_as_date", 
                            unit_var = "state",
                            policy_var = "time_first_trt_date", 
                            data = s7, 
                            max_time_to = 10e7)

s7_bm3$estimate
```

The aggregated effect estimate equals `r s7_bm2$estimate`, equivalent to the 
**Causal Effect B**.

## Summary

When the treatment effect is staggered and heterogeneous, the TWFE estimate aggregates
the heterogeneous effects using weights that are calculated by the model and
not concordant with weights researchers would intuitively choose. In this setting,
it seems sensible to estimate a separate effect for each adoption cohort, or to 
aggregate the estimates using one of the weighting schemes outlined in Section 
X.X.

`r colorize("stopped here", "green")`

`r colorize("stopped here", "need to add sun abraham")`

**Recommendation:** When there are multiple time periods, start by 
estimating a dynamic effect to see if the model supports its presence (i.e., 
does the effect change over time or is it constant?). If the effect appears 
dynamic, this is important because it is indicative of how the policy operates 
after being rolled out. If there is no strong support for a dynamic effect, then
consider estimating the effects separately by timing of introduction if that is 
sensible for the policy change under study (i.e., if there are only a few 
separate time points), or estimating one overall summary parameter.
